---
title: "Introduction"
author: "Boyko Amarov"
date: "3/14/2022"
output: html_document
---

Optional: more on R Markdown: https://rmarkdown.rstudio.com/lesson-1.html

Optional: a base R cheatsheet: https://www.datacamp.com/cheat-sheet/getting-started-r

## Arithmetic operations

Shortcut to insert a code chunk: Ctrl-Alt-i
Shortcut to run a single line of code: Ctrl-Enter
Shortcut to run the whole chunk: Ctrl-Shift-Enter

```{r}
1 + 4
3 - 2
2 * 8
2 / 8
2^4
```

## Assignment

Shortcut for the assignment operator: Alt - (minus)
```{r}
y <- 34
y - 40
```

## Vectors

```{r}
x <- c(1, 4)
```


```{r}
## Length, average, sum of a numeric vector
mean(x)
sum(x)
length(x)
```


```{r}
## Documentation
?mean
```

## Text

```{r}
z <- "Hello, world!"
```


## Logical values

```{r}
2 < 5
2 > 5
"Text 2" == "Text 2"
```

```{r}
z == "Text 2"
```

## Data

```{r}
library(tidyverse)
```

## Manual

```{r}
dt <- tibble(
  id = 1:5,
  y = c(2, 2.5, 3, 8, 12)
)
dt1 <- tribble(
  ~id, ~y,
  1, 2,
  2, 2.5,
  3, 3,
  4, 8,
  5, 12
)
dt
dt1
```

## Import from text files

```{r}
earnings <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```

- _height_ (numeric): Height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): Weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1: Male, 0: Female
- _earn_ (numeric): Annual income in USD
- _earnk_ (numeric): Annual income in 1,000 USD
- _ethnicity_ (character): Ethnicity
- _age_ (numeric): Age


```{r}
earnings <- mutate(
  earnings,
  height_cm = 2.54 * height,
  weight_kg = 0.45 * weight
)
earnings1 <- select(earnings, height_cm, weight_kg)
```

## Pipes

```{r}
earnings1 <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  ) %>%
  select(height_cm, weight_kg)
```


## Data summary

### Location

```{r}
mean(earnings1$height_cm)
## Or using a pipe
earnings1$height_cm %>% mean()
```


$$
x = (x_1, x_2,\ldots, x_n)\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^{n} x_i
$$

$n = 1816$.


### Variability

```{r}
max(earnings$height_cm)
min(earnings$height_cm)
```

```{r}
max(earnings$height_cm) - min(earnings$height_cm)
```

$$
x = (x_1, \ldots, x_n)\\
S^2_x = \frac{1}{n - 1} \sum_{i = 1}^{n}(x_i - \bar{x})^2: \quad \text{variance}\\
S_x = \sqrt{S^2_x} \quad \text{standard deviation}
$$

```{r}
var(earnings$height_cm)
```

```{r}
sqrt(var(earnings$height_cm))
```

```{r}
sd(earnings$height_cm)
```

```{r}
## Basic summaries for the whole tibble
earnings %>% skimr::skim()
```

## Visualizations

Density

```{r}
earnings %>%
  ggplot(aes(x = height)) +
  geom_density() +
  geom_histogram(aes(y = ..density..), alpha = 0.3)
```

Group comparisons

```{r}
earnings %>%
  ggplot(aes(x = height, color = ethnicity)) +
  geom_density()
```

## Boxplots

```{r}
earnings %>%
  ggplot(aes(x = height)) +
  geom_boxplot()
```

Group comparisons

```{r}
earnings %>%
  ggplot(aes(x = height, y = ethnicity)) +
  geom_boxplot()
```

Associations
```{r}
earnings1 %>%
  ggplot(aes(x = weight_kg, y = height_cm)) +
  geom_point(position = "jitter") +
  geom_smooth(method = "lm") +
  geom_smooth(color = "firebrick3")
```
