# Introduction to R

## Arithmetic Operations

Shortcut to insert a code chunk: Ctrl-Alt-i
Shortcut to run a single line of code: Ctrl-Enter
Shortcut to run the whole chunk: Ctrl-Shift-Enter

```{r}
1 + 4
3 - 2
2 * 8
2 / 8
2^4
```

## Assignment

Shortcut for the assignment operator: Alt - (minus)
```{r}
y <- 34
y - 40
```

## Vectors

```{r}
x <- c(1, 4)
```


```{r}
## Length, average, sum of a numeric vector
mean(x)
sum(x)
length(x)
```


```{r}
## Documentation
?mean
```

## Character values

```{r}
z <- "Hello, world!"
```


## Logical values and logical operators

```{r}
2 < 5
2 > 5
"Text 2" == "Text 2"
```

```{r}
z == "Text 2"
```

## Data

```{r}
library(tidyverse)
```


```{r}
dt <- tibble(
  id = 1:5,
  y = c(2, 2.5, 3, 8, 12)
)
dt
```

```{r}
earnings <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```

- _height_ (numeric): Height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): Weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1: Male, 0: Female
- _earn_ (numeric): Annual income in USD
- _earnk_ (numeric): Annual income in 1,000 USD
- _ethnicity_ (character): Ethnicity
- _age_ (numeric): Age


```{r}
earnings <- mutate(
  earnings,
  height_cm = 2.54 * height,
  weight_kg = 0.45 * weight
)
earnings1 <- select(earnings, height_cm, weight_kg)
```

The same code can be rewritten in a more convenient way using [pipes](https://r4ds.had.co.nz/pipes.html).

```{r}
earnings1 <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  ) %>%
  select(height_cm, weight_kg)
```

Note that the object holding the original data is unaffected by mutate and select.


## Basic data summaries

### Location

$$
x = (x_1, x_2,\ldots, x_n)\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^{n} x_i
$$

$n = 1816$.

```{r}
mean(earnings1$height_cm)
mean(earnings1$weight_kg, na.rm = TRUE)
```

```{r}
max(earnings1$height_cm) - min(earnings1$height_cm)
```

```{r}
var(c(1, 3))
```




```{r}
table(earnings$ethnicity)
```

```{r}
earnings %>%
  ggplot(aes(y = ethnicity)) +
  geom_bar()
```

### Variability

```{r}
max(earnings$height_cm)
min(earnings$height_cm)
```

```{r}
max(earnings$height_cm) - min(earnings$height_cm)
```

$$
x = (x_1, \ldots, x_n)\\
S^2_x = \frac{1}{n - 1} \sum_{i = 1}^{n}(x_i - \bar{x})^2: \quad \text{variance}\\
S_x = \sqrt{S^2_x} \quad \text{standard deviation}
$$
```{r}
x <- c(1, 3)
((1 - 2)^2 + (3 - 2)^2) / 1
var(x)
sd(x)
```


```{r}
var(earnings$height_cm)
```

```{r}
sqrt(var(earnings$height_cm))
```

```{r}
sd(earnings$height_cm)
```

```{r}
## Basic summaries for the whole tibble
earnings %>% skimr::skim()
```

```{r}
table(earnings$ethnicity)
```


## Visualizations

Histogram

```{r}
earnings %>%
  ggplot(aes(x = height)) +
  geom_histogram()
```

A smooth density plot is an alternative way to visualize the distribution of a variable.

```{r}
earnings %>%
  ggplot(aes(x = height)) +
  geom_density()
```

The boxplot shows the median and the 25-th and 75-th percentiles (the box). The whiskers in the plot stretch to the
minimum or the maximum observed value, unless there are extreme observations that are shown as single dots.

```{r}
earnings %>%
  ggplot(aes(x = height)) +
  geom_boxplot()
```

Group comparisons

```{r}
earnings %>%
  ggplot(aes(x = height, y = ethnicity)) +
  geom_boxplot()
```

The scatterplot will be our primary tool in studying associations between variables. It represents each
observation as a point in a coordinate system defined by the variables that we would like to study.

```{r}
earnings1 %>%
  ggplot(aes(x = weight_kg, y = height_cm)) +
  geom_point(position = "jitter", alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(
    x = "Weight (kg)",
    y = "Height (cm)"
  )
```

```{r}
summary(lm(height_cm ~ weight_kg, data = earnings1))
```
